Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-14 demystify.tex
updates

\section{Who is this document for?}

While the information presented here can be useful for all students interested in applying to a CS PhD program in the US, this guideline is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (similarly, for domestic students from universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultures and experiences might already know most of the information presented here.
My goal is to level the playing field by providing information that is not readily available to less privileged students.  I also hope to encourage more students from these countries to apply to CS PhD programs in the US.

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that most large schools, including GMU, have separate adcoms for MS programs.

@@

\paragraph{Why CS depts do not waive the application fee?}  Some universities do waive, e.g., Rice for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).
However, obtaining letters from famous people
can be challenging for international students, who might not have many interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) who know you well enough to talk about \emph{your research experience and capabilities}.

@@

If you do not waive your right, then the letter writer might refuse to write for you or write a generic letter that does not help your case.  Reviewers also might raise concerns about a letter that is not waived, e.g., if you do not trust your letter writers, then you should find someone else to write for you. In short, it's a standard practice and a way of keeping things straightforward and respectful in the whole recommendation game.

@@

\section{Graduate Assistantship (TA/RA)}\label{sec:ta-ra}
The most common type of funding is \textbf{graduate assistantship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also give significant discount or pay insurance for spouse/children.

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have extra activity fee or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

\paragraph{Low Stipend?} Students often complain their stipend being too low. However, in most cases it is actually not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At GMU, and most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

TA is common in the beginning when you haven't found an adviser who would pay you RA. It is also common to sandwich between TA and RA (e.g., when your prof. does not have sufficient funding or you want to try get TA experience).

@@

\begin{commentbox}[Vu:]
  I would strongly advocate for a non-STEM student who shows that they have a strong drive for CS by studying core CS knowledge through various channels (e.g., self-study through online courses, projects, etc.).  I have seen many students with non-CS background who are very successful in CS PhD.  I also have seen many students with CS background who are not successful in CS PhD.  So it is not about your background, it is about your drive and passion for CS research.
\end{commentbox}

@@

If you have an MS then some course work \emph{might be} transferred for course credits, which save a bit of time. But overall don't count on this, especially if your MS is not from the US.

@@

The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job.
In many cases you might take a summer or two off to do internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

\emph{MS requirement and PhD duration}:  as mentioned in \autoref{sec:msrequirement} and \autoref{sec:time}, CS PhD programs in the US do not require MS degree.  In contrast, many other countries require having an MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

@@

Furthermore, avoid using Mr. or Mrs., or the professor's first name if you're not acquainted with them yet.  As you become more familiar with your prof and depending on their preferences, you may transition to addressing them by their first name.
For example, I prefer that my students and colleague call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

PhD students often ask why their salary is so low compared to ludicrous grants their advisers get. They also wonder why their offer letters sometimes show that their benefits higher than what they actually receive (i.e., stipend).

@@

At the end, the total budget comes out to be \$72K/year to support a PhD student. \emph{The summary is that over your 5--6 years of your PhD, you cost about \$400K, and while your stipend is X, your adviser probably pays 2X for you}. But of course the nicest thing is that you do not have to pay for any of this! You get to gain the knowledge, do research, travel, (and don't even have to buy any research or computer equipments as mentioned in \autoref{sec:buying-equipment}) and also get paid!

@@

For start date or TA assignment (e.g., TA'ing a specific course), it is uncommon but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely non-negotiable.  However, you can ask for things such as computer equipment (\autoref{sec:buying-equipment}), books, (and sometimes even furniture), and a conference travel budget.

\section{Will I be miserable during my a PhD?}\label{sec:happy}
There are many stories on how students are mistreated, stressed, and miserable. Issues including bad relationships with professors, conflicts with co-authors and lab mates, feeling discriminated (e.g., because you're an international student), etc \emph{do} exist, and it is good to be aware of those.  However, in reality there are many good mentors, supportive lab mates and department, and so on.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

Most of what is written in this document is applicable to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of.

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned above, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

\paragraph{Fee Waiving} Some school might offer application fee waiver for domestic students.  You should check with the school you're applying to.

@@

 This section provides some guidance on how to gain research experience, e.g., as an undergraduate student.  It also gives advice to students from small colleges where research opportunities are limited.

\paragraph{Locally} Start looking for research opportunities at your own institution.
If you did well or liked a class, you can check with the professor of that class for research opportunities.
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you still should contact them and ask.

@@

\paragraph{Online Research Communities and Open Source Contributions} Places such as Github offer many opportunities for research, even though they might not appear so (e.g., they do not explicitly mention research).
In many cases, professors or research labs put their projects on Github. For example, my research group has a many open-source projects that undergrads can contribute to (see \url{https://github.com/dynaroars/}).

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, you might be able to get a LoR from the project maintainer, and you should write about this experience in your SOP (\autoref{sec:research-statement}.)

@@

\paragraph{Checking your application status and Follow-up emails} If you have interviewed and not heard back from a professor after a few weeks or especially around the time when universities send out their admission decisions (around late Feb-- mid Mar), you can email to check.  You can follow up the interview invitation and say: \emph{``Thanks for chatting with me. I am very excited about the opportunity to work with you.  Could you please let me know if you have made a decision or if you need more information from me?''}.  If you have new updates, e.g., new publications or new fellowship awards, or even new offers from other professors or schools, you can also mention that.

Profs. are often very busy, especially during admission time when they have to review many applications, interview many students, and make decisions.  They might not have time to respond to every email.  If you do not hear back after a week, you can send another email to check again.  If you still do not hear back, you can assume that you are not selected.

>>>>

\section*{Who is this document for?}

The information presented here can be useful for all students, but this guideline is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (similarly, for domestic students from universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultures and experiences might already know most of the information presented here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I also hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

The document might also not sufficient for students aiming for top-10 schools, which might have different requirements\footnote{e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU}.  However, the information here should be useful for students aiming for other good schools, e.g., R1 institutions in the US. It is definitely useful for students aiming for GMU :)

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that most large schools, including GMU, have separate adcoms for MS programs.

@@

\paragraph{Why CS depts do not waive the application fee?}  Some universities do waive, e.g., Rice does that for PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).

However, obtaining letters from famous people
can be challenging for international students, who might not have many interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) who know you well enough to talk about \emph{your research experience and capabilities}.

@@

If you do not waive your right,  the letter writer might refuse to write for you or write a generic letter that does not help your case.  Reviewers also might raise concerns about a letter that is not waived, e.g., if you do not trust your letter writers, then you should find someone else to write for you. In short, it's a standard practice and a way of keeping things straightforward and respectful in the whole recommendation game.

@@

\section{Graduate Assistantship (TA/RA)}\label{sec:ta-ra}
The most common type of funding is \textbf{graduate assistantship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must-have in the US), and most importantly, your stipend (i.e., your salary). Some universities also give significant discounts or pay insurance for spouses/children.

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

\paragraph{Low Stipend?} Students often complain their stipend being too low. However, in most cases it is actually not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their families with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for a 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At GMU, and most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

TA is common in the beginning when you haven't found an adviser who would pay you RA. It is also common to sandwich between TA and RA (e.g., when your prof. does not have sufficient funding or you want to try the TA experience).

@@

\begin{commentbox}[Vu:]
  I would advocate for a non-STEM student who shows that they have a strong drive for CS by studying core CS knowledge through various channels (e.g., self-study through online courses, projects, etc.).  I have seen many students with non-CS background who are very successful in CS PhD.  I also have seen many students with CS background who are not successful in CS PhD.  So it is not about your background, it is about your drive and passion for CS research.
\end{commentbox}

@@

If you have an MS then some course work \emph{might be} transferred for course credits, which saves a bit of time. But overall don't count on this, especially if your MS is not from the US.

@@

The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

\emph{MS requirement and PhD duration}:  as mentioned in \autoref{sec:msrequirement} and \autoref{sec:time}, CS PhD programs in the US do not require an MS degree.  In contrast, many other countries require having an MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

@@

Furthermore, avoid using Mr. or Mrs., or the professor's first name if you're not acquainted with them yet.  As you become more familiar with your prof and depending on their preferences, you may transition to addressing them by their first name.
For example, I prefer that my students and colleagues call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

PhD students often ask why their salary is so low compared to ludicrous grants their advisers get. They also wonder why their offer letters sometimes show that their benefits are higher than what they receive (i.e., stipend).

@@

In the end, the total budget comes out to be \$72K/year to support a PhD student. \emph{The summary is that over your 5--6 years of your PhD, you cost about \$400K, and while your stipend is X, your adviser probably pays 2X for you}. But of course, the nicest thing is that you do not have to pay for any of this! You get to gain the knowledge, do research, travel, (and don't even have to buy any research or computer equipment as mentioned in \autoref{sec:buying-equipment}) and also get paid!

@@

For start date or TA assignment (e.g., TA'ing a specific course), it is uncommon, but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely non-negotiable.  However, you can ask for things such as computer equipment (\autoref{sec:buying-equipment}), books, (and sometimes even furniture), and a conference travel budget.

\section{Will I be miserable during my a PhD?}\label{sec:happy}
There are many stories of how students are mistreated, stressed, and miserable. Issues including bad relationships with professors, conflicts with co-authors and lab mates, feeling discriminated against (e.g., because you're an international student), \emph{do} exist, and it is good to be aware of those.  However, in reality, there are many good mentors, supportive lab mates and department, and so on.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

Most of what is written in this document applies to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of.

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned above, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

\paragraph{Fee Waiving} Some schools might offer application fee waivers for domestic students.  You should check with the school you're applying to.

@@

 This section provides some guidance on how to gain research experience, e.g., as an undergraduate student.  It also advises students from small colleges where research opportunities are limited.

\paragraph{Locally} Start looking for research opportunities at your institution.
If you did well or liked a class, you can check with the professor of that class for research opportunities.
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you still should contact them and ask.

@@

\paragraph{Online Research Communities and Open Source Contributions} Places such as GitHub offer many research opportunities, even though they might not appear so (e.g., they do not explicitly mention research).
In many cases, professors or research labs put their projects on GitHub. For example, my research group has many open-source projects that undergrads can contribute to (see \url{https://github.com/dynaroars/}).

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, but you might be able to get a LoR from the project maintainer, and you should write about this experience in your SOP (\autoref{sec:research-statement}.)

@@

\paragraph{Checking your application status and Follow-up emails} If you have interviewed and not heard back from a professor after a few weeks or especially around the time when universities send out their admission decisions (around late Feb-- mid-Mar), you can email to check.  You can follow up the interview invitation and say: \emph{``Thanks for chatting with me. I am very excited about the opportunity to work with you.  Could you please let me know if you have made a decision or if you need more information from me?''}.  If you have new updates, e.g., new publications or new fellowship awards, or even new offers from other professors or schools, you can also mention that.

Profs. are often very busy, especially during admission time when they have to many reviews and interviews.  They might not have time to respond to every email.  If you do not hear back after a week, you can send another email to check again.  If you still do not hear back, you can assume that you are not selected.
